Normalization by Evaluation for Sized Dependent Types

### 2017-06-11 nbe/icfp17.tex
[ ICFP 17 ] Added sections 6 and 7 to short version.

\LONGVERSION{ % BEGIN LONGVERSION
\begin{lemma}[Properties of the logical relation for substitution]
  \label{lem:wklrsub}
  \bla \\
  Let $\LRSub \Gamma \sigma \tau \Delta \eta \rho$. Then:
  \begin{enumerate}
  \item Well-typedness: $\Gamma \der \sigma : \Delta$
   and $\Gamma \der \sigma = \sigma \eeq \tau : \Delta$
   [which in turn implies $\Gamma \der \tau : \Delta$].
  \item Weakening:
  If\/ $\xi : \Gamma' \leq \Gamma$ then
  $\LRSub{\Gamma'}{\sigma\xi}{\tau\xi}{\Delta}\eta\rho$.
  \item Resurrection:
  $\LRS {\resurrect\Gamma} \tau {\resurrect\Delta} \rho$.
  % $\LRSub {\resurrect\Gamma} \tau \tau {\resurrect\Delta} \rho \rho$.
  \item \label{it:sizesub} Size substitution:
  If\/ $\resurrect\Delta \der a : \Size$ then $a\eta \in \SIZE$.
  \end{enumerate}
\end{lemma}
} % END LONGVERSION
\LONGVERSION{
\begin{proof}
  For part (\ref{it:sizesub}), the only interesting case $a = \ind i + o$ can be proved by observing that $\eta(i) \in \SIZE$.
\end{proof}
} % END LONGVERSION
% \begin{lemma}[Typing of constructed data]
% \label{lem:invdata} \bla
% If\/ $\Gamma \der c : \Nat b$ then $\resurrect\Gamma \der b : \Size$ and $b > 0$.
% \end{lemma}

@@

\LONGVERSION{
\input{sec-asub}
} % END LONG VERSION
\SHORTVERSION{
Subtyping is decidable, for instance by full normalization
and comparison of normal forms
\cite{fridlenderPagano:tlca13}.
Alternatively, we can define the usual incremental subtyping algorithm
based on weak head evaluation, which checks equality of neutral types
by NbE.  For details, see the long version of this article.
} % END SHORTVERSION

@@

\LONGVERSION{
\input{sec-tycheck}
} % END LONG VERSION
\SHORTVERSION{
It is now straightforward to define a bidirectional type-checking
algorithm which decides typing for normal forms, and succeeds for
those that rely only on the following restricted rule for irrelevant
size application:
\[
  \ru{\Gamma \ders t : \forall T \qquad
      \resurrect\Gamma \der a : \Size
    }{\Gamma \ders t\, \ann a : T[a]}
\]
For lack of space, we delegate definition and soundness proof to the
long version of this article.
} % END SHORTVERSION

>>>>

\LONGVERSION{ % BEGIN LONGVERSION
\begin{lemma}[Properties of the logical relation for substitutions]
  \label{lem:wklrsub}
  \bla \\
  Let $\LRSub \Gamma \sigma \tau \Delta \eta \rho$. Then:
  \begin{enumerate}
  \item Well-typedness: $\Gamma \der \sigma : \Delta$
   and $\Gamma \der \sigma = \sigma \eeq \tau : \Delta$
   [which in turn implies $\Gamma \der \tau : \Delta$].
  \item Weakening:
  If\/ $\xi : \Gamma' \leq \Gamma$ then
  $\LRSub{\Gamma'}{\sigma\xi}{\tau\xi}{\Delta}\eta\rho$.
  \item Resurrection:
  $\LRS {\resurrect\Gamma} \tau {\resurrect\Delta} \rho$.
  % $\LRSub {\resurrect\Gamma} \tau \tau {\resurrect\Delta} \rho \rho$.
  \item \label{it:sizesub} Size substitution:
  If\/ $\resurrect\Delta \der a : \Size$ then $a\eta \in \SIZE$.
  \end{enumerate}
\end{lemma}
} % END LONGVERSION
\LONGVERSION{
\begin{proof}
  For part (\ref{it:sizesub}), the only interesting case $a = \ind i + o$ can be proved by observing that $\eta(i) \in \SIZE$.
\end{proof}
} % END LONGVERSION
% \begin{lemma}[Typing of constructed data]
% \label{lem:invdata} \bla
% If\/ $\Gamma \der c : \Nat b$ then $\resurrect\Gamma \der b : \Size$ and $b > 0$.
% \end{lemma}

@@

\input{sec-asub}
% \LONGVERSION{
% \input{sec-asub}
% } % END LONG VERSION
% \SHORTVERSION{
% Subtyping is decidable, for instance by full normalization
% and comparison of normal forms
% \cite{fridlenderPagano:tlca13}.
% Alternatively, we can define the usual incremental subtyping algorithm
% based on weak head evaluation, which checks equality of neutral types
% by NbE.  For details, see the long version of this article.
% } % END SHORTVERSION

@@

\input{sec-tycheck}
% \LONGVERSION{
% \input{sec-tycheck}
% } % END LONG VERSION
% \SHORTVERSION{
% It is now straightforward to define a bidirectional type-checking
% algorithm which decides typing for normal forms, and succeeds for
% those that rely only on the following restricted rule for irrelevant
% size application:
% \[
%   \ru{\Gamma \ders t : \forall T \qquad
%       \resurrect\Gamma \der a : \Size
%     }{\Gamma \ders t\, \ann a : T[a]}
% \]
% For lack of space, we delegate definition and soundness proof to the
% long version of this article.
% } % END SHORTVERSION

### 2017-06-11 nbe/sec-asub.tex
[ ICFP 17 ] Figurized algorithmic subtyping.
[ ICFP 17 ] Added sections 6 and 7 to short version.

Algorithmic subtyping \fbox{$\Gamma \der T \asub T'$}.
\begin{gather*}
  \ru{T \evalsto n \qquad
      T' \evalsto n' \qquad
      \Nbe \Gamma n \nfeq \Nbe \Gamma n'
    }{\Gamma \der T \asub T'}
\qquad
  \rux{T \evalsto \Set[\ell] \qquad
      T' \evalsto \Set[\ell']
    }{\Gamma \der T \asub T'
    }{\ell \leq \ell'}
\\[2ex]
  \rux{T \evalsto \Nat a \qquad
      T' \evalsto \Nat a'
    }{\Gamma \der T \asub T'
    }{a \leq a'}
\qquad
  \ru{T_1' \evalsto \epiT\Size{T_1} \quad \
      T_2' \evalsto \epiT\Size{T_2} \quad \
      \cext \Gamma {\Size} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\\[2ex]
  \ru{T_1' \evalsto \piT{U_1}{T_1} \qquad
      T_2' \evalsto \piT{U_2}{T_2} \qquad
      \Gamma \der U_2 \asub U_1 \qquad
      \cext \Gamma {U_2} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\end{gather*}

\begin{lemma}[Soundness of algorithmic subtyping]
  \label{lem:asubsound}
  If\/ $\Gamma \der T \asub T'$ then $\Gamma \der T \leq T'$.
\end{lemma}
\begin{proof}
  By induction on $\Gamma \der T \asub T'$, soundness of NbE, and subject reduction (Thm.~\ref{thm:sr}).
\end{proof}

\begin{lemma}[Semantic subtyping implies algorithmic subtyping]
\label{lem:lrasub}
\bla
   If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
\end{lemma}
\begin{proof*}
By induction on % $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
$\LRTS \Gamma {T_1'} {A_1'} {s_1}$ and $\LRTS \Gamma {T_2'} {A_2'} {s_2}$
and cases on $A_1' \leq A_2' \in \TYPE$
we prove $\Gamma \der T_1' \asub T_2'$.

@@

\nextcase $A_i' \evalsto \piT {A_i} {B_i}$ and  $T_i' \evalsto \piT{U_i}{T_i}$.
From $\LRTS \Gamma {U_i} {A_i} {s_i}$ and $A_2 \leq A_1 \in \TYPE$ we get $\Gamma \der U_2 \asub U_1$
by induction hypothesis on $A_i \in s_i$.
Let $a := {\up{A_2} \var\Gamma}$.  With $\LR {\cext \Gamma {U_2}} {\ind 0} {(U_2\slift)} a {A_2}$ we get
$\LRTS {\cext \Gamma {U_2}} {T_i} {B_i[a]} s$ and $B_1[a] \leq B_2[a] \in \TYPE$.  Thus by induction hypothesis on $B_i[a] \in s_i$ we obtain $\cext \Gamma {U_2} \der T_1 \leq T_2$, together $\Gamma \der T_1' \asub T_2'$.
\qed
\end{caselist}
\end{proof*}

>>>>

Fig~\ref{fig:asub} defines an incremental subtyping algorithm
\fbox{$\Gamma \der T \asub T'$}.
Neutral types are subtypes iff they are equal, which is checked using NbE.
\begin{figure}[htbp]
\hrulefill
\begin{gather*}
  \ru{T \evalsto n \qquad
      T' \evalsto n' \qquad
      \Nbe \Gamma n \nfeq \Nbe \Gamma n'
    }{\Gamma \der T \asub T'}
\qquad
  \rux{T \evalsto \Set[\ell] \qquad
      T' \evalsto \Set[\ell']
    }{\Gamma \der T \asub T'
    }{\ell \leq \ell'}
\\[2ex]
  \rux{T \evalsto \Nat a \qquad
      T' \evalsto \Nat a'
    }{\Gamma \der T \asub T'
    }{a \leq a'}
\qquad
  \ru{T_1' \evalsto \epiT\Size{T_1} \quad \
      T_2' \evalsto \epiT\Size{T_2} \quad \
      \cext \Gamma {\Size} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\\[2ex]
  \ru{T_1' \evalsto \piT{U_1}{T_1} \qquad
      T_2' \evalsto \piT{U_2}{T_2} \qquad
      \Gamma \der U_2 \asub U_1 \qquad
      \cext \Gamma {U_2} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\end{gather*}
\hrulefill
\caption{Algorithmic subtyping \fbox{$\Gamma \der T \asub T'$}.}
\label{fig:asub}
\end{figure}

\begin{lemma}[Soundness of algorithmic subtyping]
  \label{lem:asubsound}
  If\/ $\Gamma \der T \asub T'$ then $\Gamma \der T \leq T'$.
\end{lemma}
\begin{proof}
  By induction on $\Gamma \der T \asub T'$, soundness of NbE, and
  subject reduction\LONGVERSION{~(Thm.~\ref{thm:sr})}.
\end{proof}

\begin{lemma}[Semantic subtyping implies algorithmic subtyping]
\label{lem:lrasub}
\bla \\
   If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
\end{lemma}
\SHORTLONG{
\begin{proof}
  By induction on $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and
  cases on $A \leq A' \in \TYPE$.
\end{proof}
}{ % LONG VERSION
\begin{proof*}
By induction on % $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
$\LRTS \Gamma {T_1'} {A_1'} {s_1}$ and $\LRTS \Gamma {T_2'} {A_2'} {s_2}$
and cases on $A_1' \leq A_2' \in \TYPE$
we prove $\Gamma \der T_1' \asub T_2'$.

@@

\nextcase $A_i' \evalsto \piT {A_i} {B_i}$ and  $T_i' \evalsto \piT{U_i}{T_i}$.
From $\LRTS \Gamma {U_i} {A_i} {s_i}$ and $A_2 \leq A_1 \in \TYPE$ we get $\Gamma \der U_2 \asub U_1$
by induction hypothesis on $A_i \in s_i$.
Let $a := {\up{A_2} \var\Gamma}$.  With $\LR {\cext \Gamma {U_2}} {\ind 0} {(U_2\slift)} a {A_2}$ we get
$\LRTS {\cext \Gamma {U_2}} {T_i} {B_i[a]} s$ and $B_1[a] \leq B_2[a] \in \TYPE$.  Thus by induction hypothesis on $B_i[a] \in s_i$ we obtain $\cext \Gamma {U_2} \der T_1 \leq T_2$, together $\Gamma \der T_1' \asub T_2'$.
\qed
\end{caselist}
\end{proof*}
} % END LONG VERSION
